The imprint of global magma oceans on exoplanet demographics

### 2023-04-09 src/tex/ms.tex
6.3.

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{2019NatAs...3..307L,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}
a
We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interace \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.

@@

As discussed above, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
%In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets studied with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
Overall, it seems feasible to derive the typical water content of rocky planets from these occurrence estimates.

@@

Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
Such direct imaging surveys, however, do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets. \todo[inline]{TIM: feel free to add something about LIFE if you see fit}

As for exoplanets missions in their implementation phase, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.

>>>>

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
%\todo[inline]{TIM: Ideas?\\
%discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3 \citep{2022AJ....164..172D}, or Kepler-138 \citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.

@@

As discussed above, key drivers from a mission design perspective are sample size, photometric precision, and the availability of planets around low-mass host stars.
We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
The \plato\ team has released an estimate on the number of exoplanets that will be characterized in the course of the main survey mission (Matuszewski, priv. comm.).
%In the FGK sample of the mission, several hundred Earth-sized (\SIrange{0.8}{1.25}{\rEarth}) planets are expected to be detected.
With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets studied with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.
The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
Overall, it seems feasible to derive the typical water content of low-mass exoplanets from these occurrence estimates.

@@

Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets. However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths \citep{2018ExA....46..543D,2021ExA...tmp..118Q}. Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light \citep{Wordsworth2022}. Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+). Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages \citep{2014ApJ...784...27L,2019A&A...621A.125B}. With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

As for exoplanets missions in their implementation phase, however, \plato\ overall remains to be the most promising mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.
